---
title: "MSDS6301_CaseStudy2"
author: "Fred Poon"
date: "12/5/2019"
output: slidy_presentation
---

```{r Libraries, echo=FALSE, results='hide'}

# install packages
install.packages("readxl")
install.packages("tidyverse")
install.packages("psych")
install.packages("class")
install.packages("caret")
install.packages("e1071")

# load packages
library(readxl)         ## this reads excel (xls, xlsx) files
library(tidyverse)      ## tidyverse  by default
library(psych)          ## using this package to dummy.code categorical variables.
library(class)          ## knn and naiveBayes
library(caret)          ## confusionMatrix
library(e1071)          ## confusionMatrix

```

```{r Data, echo=TRUE, results='hide'}

#load employee management data sets
#employees dataset - holds attrition and income values
data = read.csv('C:/Users/poonf/OneDrive - Southern Methodist University/6306 doing data science/project 2/CaseStudy2-data.csv', header = TRUE)

#the following are variables with not enough background information
#StandardHours EmployeeNumber EmployeeCount Over18
#we will remove these columns from dataset
data <- select(data, -StandardHours, -EmployeeNumber, -EmployeeCount, -Over18)

#attrition test set
attrition = read.csv('C:/Users/poonf/OneDrive - Southern Methodist University/6306 doing data science/project 2/CaseStudy2CompSet No Attrition.csv', header = TRUE)

#income test set
income = read_xlsx('C:/Users/poonf/OneDrive - Southern Methodist University/6306 doing data science/project 2/CaseStudy2CompSet No Salary.xlsx', col_names = TRUE)

```

### Information gathered from Excel column names.
StandardHours = 80... does this imply bi-weekly payroll?
EmployeeCount = 1... Not sure what this means..
EmployeeNumber - what significance does this have?
Over18 - not necessary since we have age

```{r Exploratory - Summary Statistics}

#out of curiousity, some summary statistics for selected columns
summary(data)

```

## Interesting statistics
Here are some variables I'd like to hone in on.
Age, Attrition, BusinessTravel, Department, DistanceFromHome, EducationField, Gender, JobRole, MaritalStatus, MonthlyIncome, NumCompaniesWorked, OverTime, PercentSalaryHike, StockOptionLevel, TotalWorkingYears, YearsAtCompany, YearsInCurrentRole, YearsSinceLastPromotion, YearswithCurrManager

```{r Exploratory 2 - Attrition vs..}

#here i wanted to compare the yes/no attrition with numerical values
#I wanted to stay away from ratings because 1- I don't know whats 1 vs 4 and 2- I don't like the rating system due to forced or lazy answers...
data %>% ggplot(aes(Attrition, Age)) + 
  geom_boxplot() + 
  ggtitle("Age vs Attrition")

data %>% ggplot(aes(Attrition, DistanceFromHome)) + 
  geom_boxplot() + 
  ggtitle("DistanceFromHome vs Attrition")

data %>% ggplot(aes(Attrition, MonthlyIncome)) + 
  geom_boxplot() + 
  ggtitle("MonthlyIncome vs Attrition")

data %>% ggplot(aes(Attrition, NumCompaniesWorked)) + 
  geom_boxplot() + 
  ggtitle("NumCompaniesWorked vs Attrition")

data %>% ggplot(aes(Attrition, PercentSalaryHike)) + 
  geom_boxplot() + 
  ggtitle("PercentSalaryHike vs Attrition")

data %>% ggplot(aes(Attrition, StockOptionLevel)) + 
  geom_boxplot() + 
  ggtitle("StockOptionLevel vs Attrition")

data %>% ggplot(aes(Attrition, TotalWorkingYears)) + 
  geom_boxplot() + 
  ggtitle("TotalWorkingYears vs Attrition")

data %>% ggplot(aes(Attrition, YearsAtCompany)) + 
  geom_boxplot() + 
  ggtitle("YearsAtCompany vs Attrition")

data %>% ggplot(aes(Attrition, YearsInCurrentRole)) + 
  geom_boxplot() + 
  ggtitle("YearsInCurrentRole vs Attrition")

data %>% ggplot(aes(Attrition, YearsSinceLastPromotion)) + 
  geom_boxplot() + 
  ggtitle("YearsSinceLastPromotion vs Attrition")

data %>% ggplot(aes(Attrition, YearsWithCurrManager)) + 
  geom_boxplot() + 
  ggtitle("YearsWithCurrManager vs Attrition")

```

From the boxplots, we can see relationships especially in the following categories:
Age, MonthlyIncome, TotalWorkingYears, YearsAtCompany, YearsInCurrentRole,

Other thoughts, Age: 
- some outliers in Age which really just tell that older folks are either retiring or moving on.

Other thoughts, YearsInCurrRole:
- probably should leave this category out. "Yes's" should heavily skew to the right due to the relationship between Years and Turnover.

### Categorical Variables
BusinessTravel, Department, Education, EducationField, Gender, JobLevel, JobRole, MaritalStatus, OverTime, StockOptionLevel,

### Rating Variables
EnvironmentSatisfaction, JobInvolvement(?), JobSatisfaction, PerformanceRating, RelationshipSatisfaction, TrainingLastYear, WorkLifeBalance

### Numerical Values
Age, DistanceFromHome, MonthlyIncome, NumCompaniesWorked, PercentSalaryHike, TotalWorkingYears, TrainingTimeLastYear, YearsAtCompany, YearsInCurrentRole, YearsSinceLastPromotion, YearswithCurrManager

```{r Exploratory 3 continued..}

#knowing what I know above, I want to scatterplot significant relationships with categories using facet_wrap()

data %>% ggplot() +
  geom_point(aes(Age, MonthlyIncome, color = Attrition)) +
  geom_smooth(aes(Age, MonthlyIncome, linetype = Attrition)) +
  facet_wrap(~Attrition) +
  ggtitle("Age vs Income by Attrition")
#makes sense, large cluster of high income individuals under "No" Attrition. 

data %>% ggplot() +
  geom_point(aes(YearsAtCompany, MonthlyIncome, color = Attrition)) +
  geom_smooth(aes(YearsAtCompany, MonthlyIncome, linetype = Attrition)) +
  facet_wrap(~Attrition) +
  ggtitle("YearsAtCompany vs MonthlyIncome by Attrition")

data %>% ggplot() +
  geom_col(aes(WorkLifeBalance, MonthlyIncome)) +
  facet_wrap(~MaritalStatus) +
  ggtitle("WorkLifeBalance vs MonthlyIncome by MaritalStatus")

data %>% ggplot() +
  geom_col(aes(WorkLifeBalance, MonthlyIncome)) +
  facet_wrap(~Gender) +
  ggtitle("WorkLifeBalance vs MonthlyIncome by Gender")

```

```{r K-NN}

#dummy.coding categories for better k-nn predictions
BusinessTravel <- as.data.frame(dummy.code(data$BusinessTravel))
Department <- as.data.frame(dummy.code(data$Department))
EducationField <- as.data.frame(dummy.code(data$EducationField))
EducationField <- rename(EducationField, HREd = `Human Resources`)
Gender <- as.data.frame(dummy.code(data$Gender))
JobRole <- as.data.frame(dummy.code(data$JobRole))
JobRole <- rename(JobRole, HRJr = `Human Resources`)
MaritalStatus <- as.data.frame(dummy.code(data$MaritalStatus))
OverTime <- as.data.frame(dummy.code(data$OverTime))
OverTime <- rename(OverTime, NoOverTime = No)
OverTime <- rename(OverTime, YesOverTime = Yes)

#combine dataset with dummy.code
data <- cbind(data, BusinessTravel, Department, EducationField, Gender, JobRole, MaritalStatus, OverTime)
#confirm structure of new data set
str(data)

#final dataset stored as data2
data2 <- data %>% 
  select(Attrition,Age,DailyRate,DistanceFromHome,Education,EnvironmentSatisfaction,HourlyRate,JobInvolvement,JobLevel,JobSatisfaction,MonthlyIncome,MonthlyRate,NumCompaniesWorked,PercentSalaryHike,PerformanceRating,RelationshipSatisfaction,StockOptionLevel,TotalWorkingYears,TrainingTimesLastYear,WorkLifeBalance,YearsAtCompany,YearsInCurrentRole,YearsSinceLastPromotion,YearsWithCurrManager,`Non-Travel`,Travel_Frequently,Travel_Rarely,`Research & Development`,Sales,HREd,`Life Sciences`,Marketing,Medical,Other,`Technical Degree`,Female,Male,`Healthcare Representative`,HRJr,`Laboratory Technician`,Manager,`Manufacturing Director`,`Research Director`,`Research Scientist`,`Sales Executive`,`Sales Representative`,Divorced,Married,Single,NoOverTime,YesOverTime)

#K-NN model
set.seed(6)
splitPerc = .80

trainIndices = sample(1:dim(data2)[1],round(splitPerc * dim(data2)[1]))
train = data2[trainIndices,]
test = data2[-trainIndices,]

#view(train)
#str(train)
#view(test)

classifications = knn(train[,c(2:51)],test[,c(2:51)],train$Attrition, prob = TRUE, k = 3)
table <- table(classifications,test$Attrition)
CM <- confusionMatrix(table)

table
CM

## Sensitivity 88%
## Specificity 14%

```

```{r}

#let's try "normalizing the data"
normalize <- function(x) {
  return ((x-min(x)) / (max(x) - min(x))) 
}

data3 <- data2

data3_n <- as.data.frame(lapply(data3[2:51],normalize))
str(data3_n)

data3_n2 <- cbind(Attrition = data3[,c(1)],data3_n)
view(data3_n2)


#K-NN on normalized  data
trainIndices2 = sample(1:dim(data3_n2)[1],round(splitPerc * dim(data3_n2)[1]))
train2 = data3_n2[trainIndices2,]
test2 = data3_n2[-trainIndices2,]

view(train2)
view(test2)

sum(is.na(train2))
sum(is.na(test2))

classifications2 = knn(train2[,c(2:51)],test2[,c(2:51)],train2$Attrition, prob = TRUE, k = 3)
table2 <- table(classifications2,test2$Attrition)
CM2 <- confusionMatrix(table)

table2
CM2


## Sensitvity >80%
## Specificity >10%

#not sure why i was expecting different results

```
## Going for NaiveBayes Model

```{r}

#Going to try NaiveBayes
trainIndices3 = sample(1:dim(data3_n2)[1],round(splitPerc * dim(data3_n2)[1]))
train3 = data3_n2[trainIndices3,]
test3 = data3_n2[-trainIndices3,]

nBmodel = naiveBayes(Attrition ~ ., train3)
table3 <- table(predict(nBmodel,test3),test3$Attrition)
CM3 <- confusionMatrix(table3)

CM3

## Sensitivity >70%
## Specificity >70%

BusinessTravel <- as.data.frame(dummy.code(attrition$BusinessTravel))
Department <- as.data.frame(dummy.code(attrition$Department))
EducationField <- as.data.frame(dummy.code(attrition$EducationField))
EducationField <- rename(EducationField, HREd = `Human Resources`)
Gender <- as.data.frame(dummy.code(attrition$Gender))
JobRole <- as.data.frame(dummy.code(attrition$JobRole))
JobRole <- rename(JobRole, HRJr = `Human Resources`)
MaritalStatus <- as.data.frame(dummy.code(attrition$MaritalStatus))
OverTime <- as.data.frame(dummy.code(attrition$OverTime))
OverTime <- rename(OverTime, NoOverTime = No)
OverTime <- rename(OverTime, YesOverTime = Yes)

attrition <- cbind(attrition, BusinessTravel, Department, EducationField, Gender, JobRole, MaritalStatus, OverTime)

attrition2 <- attrition %>% 
  select(Age,DailyRate,DistanceFromHome,Education,EnvironmentSatisfaction,HourlyRate,JobInvolvement,JobLevel,JobSatisfaction,MonthlyIncome,MonthlyRate,NumCompaniesWorked,PercentSalaryHike,PerformanceRating,RelationshipSatisfaction,StockOptionLevel,TotalWorkingYears,TrainingTimesLastYear,WorkLifeBalance,YearsAtCompany,YearsInCurrentRole,YearsSinceLastPromotion,YearsWithCurrManager,`Non-Travel`,Travel_Frequently,Travel_Rarely,`Research & Development`,Sales,HREd,`Life Sciences`,Marketing,Medical,Other,`Technical Degree`,Female,Male,`Healthcare Representative`,HRJr,`Laboratory Technician`,Manager,`Manufacturing Director`,`Research Director`,`Research Scientist`,`Sales Executive`,`Sales Representative`,Divorced,Married,Single,NoOverTime,YesOverTime)

view(attrition2)
str(attrition2)
str(attrition)

##export to 
attrition_exp <- predict(nBmodel,attrition2)
exp = cbind.data.frame(attrition$ID,attrition_exp)
colnames(exp) <- c("ID","Attrition")
write.csv(exp,"C:/Users/poonf/OneDrive - Southern Methodist University/6306 doing data science/project 2/Case2PredictionsPOON_Attrition.csv")

```

```{r Monthly Income}

#I'm going to use the same dataset from training for Attrition to predict MonthlyIncome, instead of NaiveBayes or KNN I'm going to use linear regression

BusinessTravel <- as.data.frame(dummy.code(income$BusinessTravel))
Department <- as.data.frame(dummy.code(income$Department))
EducationField <- as.data.frame(dummy.code(income$EducationField))
EducationField <- rename(EducationField, HREd = `Human Resources`)
Gender <- as.data.frame(dummy.code(income$Gender))
JobRole <- as.data.frame(dummy.code(income$JobRole))
JobRole <- rename(JobRole, HRJr = `Human Resources`)
MaritalStatus <- as.data.frame(dummy.code(income$MaritalStatus))
OverTime <- as.data.frame(dummy.code(income$OverTime))
OverTime <- rename(OverTime, NoOverTime = No)
OverTime <- rename(OverTime, YesOverTime = Yes)

income <- cbind(income, BusinessTravel, Department, EducationField, Gender, JobRole, MaritalStatus, OverTime)

income2 <- income %>% 
  select(Attrition,Age,DailyRate,DistanceFromHome,Education,EnvironmentSatisfaction,HourlyRate,JobInvolvement,JobLevel,JobSatisfaction,MonthlyRate,NumCompaniesWorked,PercentSalaryHike,PerformanceRating,RelationshipSatisfaction,StockOptionLevel,TotalWorkingYears,TrainingTimesLastYear,WorkLifeBalance,YearsAtCompany,YearsInCurrentRole,YearsSinceLastPromotion,YearsWithCurrManager,`Non-Travel`,Travel_Frequently,Travel_Rarely,`Research & Development`,Sales,HREd,`Life Sciences`,Marketing,Medical,Other,`Technical Degree`,Female,Male,`Healthcare Representative`,HRJr,`Laboratory Technician`,Manager,`Manufacturing Director`,`Research Director`,`Research Scientist`,`Sales Executive`,`Sales Representative`,Divorced,Married,Single,NoOverTime,YesOverTime)

lmodel <- lm(MonthlyIncome ~ ., data3)

income_exp <- predict(lmodel,income2)
exp_i = cbind.data.frame(income$ID,income_exp)
colnames(exp_i) <-c("ID","MonthlyIncome")
write.csv(exp_i,"C:/Users/poonf/OneDrive - Southern Methodist University/6306 doing data science/project 2/Case2PredictionsPOON_Salary.csv")

```









